---
title: "R Notebook"
output: html_notebook
---


## Load the murder data for Arizona

Don't forget to load up the tidyverse using the library() command. 
Then, you just have to tell R where to find the document. You can either copy it into the directory of your project, give it the full path, or use it as a URL. Here's an example using it as a URL -- just note the file=url(...) instead of file="..." command.

I've fixed up the file a little for you, including codes for some of the categories and lower case names. That's why it's v2


```{r}
library(tidyverse)

#load ("murder_data_v2.Rdata")
load (url("https://cronkitedata.github.io/cronkite-docs/r-stats/murder_data_v2.RData"))

#print out the structure of the dataset
str(murder_data_v2)

```

There is a new kind of variable, called a "Factor". This is used when there are categories that won't change. For example, the sex is either "Female", "Male" or "Unknown" .  For now, don't worry about it -- the tidyverse is very forgiving of treating them the same as character fields. 

To get the choices of a character or factor variable, use  a new verb -- "distinct"


```{r}

murder_data %>% select ( Weapon) %>%
    mutate (weapon_code = as.integer(Weapon)) %>%
    group_by (weapon_code, Weapon) %>%
    summarise (n())


```

```{r}

distinct(murder_data, VicSex)

```

```{r}

#just rename the old data frame
murder_data_v2 <- new_murder_data
save (murder_data_v2, file="~/Documents/Github/cronkite-docs/docs/r-stats/murder_data_v2.RData")

load ("~/Documents/Github/cronkite-docs/docs/r-stats/murder_data_v2.RData")

```

