---
title: "Merging tables"
output: 
    html_document: 
       df_print: paged
    html_notebook: default
---


(With thanks to Andrew Ba Tran, whose [R for Journalists course](https://learn.r-journalism.com/en/) this is based upon.)

This notebook will go over grouping and merging in R, and will give you a bonus: How to get data from a website using a simple table command. 

We have a new version of our dataset, this time a little larger with Washington, Oregon and Idaho added to our Arizona data. I've gotten rid of some of the traps -- all of the field names are now lower cased, and I removed some of the problems with extra spaces. 

We're going to see if we can replicate an analysis of the Green River Killer.  This is one claim from the Serial Killer Detective: 

<blockquote>
Ridgway’s slayings began in 1982, when young runaways and prostitutes began disappearing from state Route 99 in south King County, Washington. He brought many of them to his home and strangled them, then left them in woodsy, remote sites. The first few bodies turned up along the now-notorious Green River.

Ridgway told investigators he killed as many as 75-80 women along Route 99 in south King County, Washington. He was convicted and received multiple life sentences.
</blockquote>




## Packages used in this tutorial


We haven't loaded an Excel spreadsheet yet -- just files from the web and a CSV. Although "readxl" is part of the tidyverse, you have to activate it separately using the library command: 

```{r}
library(tidyverse)
library(readxl)


#this is the new version
load("murders.Rda")

```

## Step 1 : We need county names! 

Your data has a field called cntyfips -- that's a common way of looking at geographic variables. It's a standard code that's used by agencies to make sure that it's always the same, no matter how a county is spelled. 

Luckily, it's easy to get data for counties through the Census and other sites. I downloa](https://socialexplorer.com), which is an easy way for ASU students to grab Census related data. 

This time, instead of importing a csv, I'll import an Excel spreadsheet called countypops.xlsx, saved in my project.  These are average county-level population estimates for the years 2013 through 2016 from the American Community Survey. 


```{r}

counties <- read_excel("countypops.xlsx")
str(counties)

```


Now we have a list of numeric FIPS codes that can be matched against the converted numeric FIPS codes in our murder data. 

## The magic merge 

One of the hardest things to do in Excel is to merge two datasets together. In R (and in other programming languages), it's not only expected that you have to do this, but it's considered best practices. 

As in other programming languages, we'll use a term called JOIN to mix together two datasets. 

This is, in journalism terms, called an "intentional join". We often use what journalists call an "enterprise join", when we are trying to find things on two lists that don't belong. We'll get to that in a later lesson. For now, our joins will follow the rules that database experts use: 

* The list of codes is unique -- there aren't any repeats. Otherwise, every match will be repeated.
* The list of codes is complete -- there aren't any missing codes. This actually isn't true in our case, but it doesn't affect our results, so we'll not worry about it. 
* The list of codes is exactly the same -- in both data type and value -- as the values we want to match. Close doesn't count.

They way you join is to add it to your list of operations the same way we did before:

```{r}
murders_with_counties <-
  murders %>%
     inner_join ( counties, by= c("cntyfips"= "fips_num")) %>%
     #let's just keep some of the variables in a different order
     #and renaming the county_name field so that it's a little shorter
     select (id, ori, agency, state, cntyfips, county=county_name, 
             solved, year, situation, weapon, 
             victim_age, victim_sex, victim_race, vic_relate_to_offender, 
             circumstance, offender_age, offender_race) 


#let's see which agencies are in which counties: 
#
murders_with_counties %>%
  group_by (state, county, agency) %>%
  summarise (murders = n())

```




# Build the query

Try building the query one step at a time in the following code block: 

What are the weapons

```{r}
 murders_with_counties %>%
  group_by(weapon) %>% 
  summarize (n())



```




What do we have to find? 

* Beginning in 1982
* southern King County, Washington
* Mainly strangled
* Victims were young runaways or prostitutes


I've introduced a new idea: str_detect(county, "King"). This  means that the code should look for upper-cased King anywhere in the county field. If I wanted to it start with King, I would use str_detect("^King")

```{r}
# type your code here 

murders_with_counties %>%
  
  filter ( year >= 1982  &
           str_detect (county, "^King") &
           state == "Washington" & 
           str_detect (weapon, "Strangulation") &
           victim_sex != "Male"
  )
             



```


## Important caveat

Of course, this isn't really finding a serial killer, and we don't know if what we found is right. But it's a way to start narrowing down what to look for when you get a tip.

## Group by and merge

Another thing you might use group_by and join is to merge once you've added up something. In this case, let's say we want to get the approximate murder rate for each county in our sample of several states. 

### Step 1: Summarise by county

In this case, we want to get a county-level data frame that is made up of the number of murders by county from 2013 through 2016, the years of our population data. 

```{r}

county_murders <- 
  murders %>%
  filter (between (year, 2013, 2016)) %>%
  group_by (cntyfips, state) %>%
  summarise (murder_ct = n())
  

```

Now merge it with the population data, and calculate a rate per 100,000 residents!

```{r}

county_murders %>%
  # I don't know why, but the names of these fields have to be in quotes
  inner_join (counties, by=c("cntyfips"="fips_num")) %>%
  select (state, county_name, cntyfips, murder_ct, total_pop_2013to16) %>%
  mutate ( murder_rate = murder_ct / total_pop_2013to16 * 100000) %>%
  arrange (desc (murder_rate))

```

Let's do it again, only this time fiter out little counties of fewer than 20,000 people. 

```{r}

county_murders %>%
  # I don't know why, but the names of these fields have to be in quotes
  inner_join (counties, by=c("cntyfips"="fips_num")) %>%
  select (state, county_name, cntyfips, murder_ct, total_pop_2013to16) %>%
  filter (total_pop_2013to16 >= 20000) %>%
  mutate ( murder_rate = murder_ct / total_pop_2013to16 * 100000) %>%
  arrange (desc (murder_rate))


```

One thing that sticks out here: Arizona only has 15 counties -- there are 87 across the four states we've looked at. But Arizona is 5 of the top 10. 





